Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
some more changes

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

A number of languages have been implemented with PyPy. The project was started
to get a better Python implementation, which inspired the name of the project
and is still the main focus of development. In addition a number of other
languages were implemented, among them a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010}, a Smalltalk VM
\cite{carl_friedrich_bolz_back_2008} and a GameBoy emulator
\cite{bruni_pygirl:_2009}.

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer by a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code, in addition
to the original backend for the Intel \emph{x86} architecture.  Examples of
additional JIT backends are the one for Intel \emph{x86-64} and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

This trace of operations is then the basis of the generated code. The trace is
optimized in some ways, and then turned into machine code. Both optimizations
and machine code generation is simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

@@

For the purpose of this paper, we are going to use a very simple object
model, that just supports an integer and a float type. The objects support only
two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
than zero. The implementation of \texttt{add} uses classical Smalltalk-like
double-dispatching. These classes could be part of the implementation of a very
simple interpreter written in RPython. The classes can be seen in
Figure~\ref{fig:objmodel}.

\begin{figure}
\begin{verbatim}
class Base(object):
    def add(self, other):
        """ add self to other """
        raise NotImplementedError("abstract base")
    def add__int(self, intother):
        """ add intother to self,
            where intother is an integer """
        raise NotImplementedError("abstract base")
    def add__float(self, floatother):
        """ add floatother to self,
            where floatother is a float """
        raise NotImplementedError("abstract base")
    def is_positive(self):
        """ returns whether self is positive """
        raise NotImplementedError("abstract base")

@@

def f(y):
    res = BoxedInteger(0)
    while y.is_positive():
        res = res.add(y).add(BoxedInteger(-100))
        y = y.add(BoxedInteger(-1))
    return res
\end{verbatim}
\caption{A Simple Object Model and an Example Function Using it}
\label{fig:objmodel}
\end{figure}

@@

To understand the problems more directly, let us consider the simple function
\texttt{f} that uses the object model (see the bottom of
Figure~\ref{fig:objmodel}).

@@

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \texttt{y} and \texttt{res} in the original function.

>>>>

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

A number of languages have been implemented with PyPy. The project was initiated
to get a better Python implementation, which inspired the name of the project
and is still the main focus of development. In addition a number of other
languages were implemented, among them a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010}, a Smalltalk VM
\cite{carl_friedrich_bolz_back_2008} and a GameBoy emulator
\cite{bruni_pygirl:_2009}.

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer using a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code, in addition
to the original backend for the Intel \emph{x86} architecture.  Examples of
additional JIT backends are the one for Intel \emph{x86-64} and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
PyPy's JIT generator generates a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

This trace of operations is then the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimizations
and machine code generation is simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

@@

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type. The objects support only
two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
than zero. The implementation of \texttt{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
\ignore{
class Base(object):
    def add(self, other):
        """ add self to other """
        raise NotImplementedError("abstract base")
    def add__int(self, intother):
        """ add intother to self,
            where intother is an integer """
        raise NotImplementedError("abstract base")
    def add__float(self, floatother):
        """ add floatother to self,
            where floatother is a float """
        raise NotImplementedError("abstract base")
    def is_positive(self):
        """ returns whether self is positive """
        raise NotImplementedError("abstract base")
        }
\begin{figure}
\begin{verbatim}
class Base(object):

@@

def f(y):
    res = BoxedInteger(0)
    while y.is_positive():
        res = res.add(y).add(BoxedInteger(-100))
        y = y.add(BoxedInteger(-1))
    return res
\end{verbatim}
\caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
%\caption{A Simple Object Model and an Example Function Using it}
\label{fig:objmodel}
\end{figure}

@@

To understand the problems more directly, let us consider the simple
interpreter function
\texttt{f} that uses the object model (see the bottom of
Figure~\ref{fig:objmodel}).

@@

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).

XXX in which language is the trace written in ? still RPython ?

The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \texttt{y} and \texttt{res} in the original function.

XXX explain set and get + int_add briefly
